Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-07 analysis/paper/paper.Rmd
Started assembling probabilistic sampling of dates for curation index plot

```{r libraries, eval = TRUE}
library(tidyverse)
library(GGally)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

@@

The analyses conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials and seasonality. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.

Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

```{r c14-modelling}
# define function to model radiocarbon dates. Only assumption is that 
# dates from a single site belong to a single group
model_rcarb <- function(data){

@@

  # Run OxCal code - wrapped this in some functions to  not print to 
  # rmarkdown
  invisible(capture.output(oxcal_exec <- executeOxcalScript(oxcal_code)))
  oxcal_read <- readOxcalOutput(oxcal_exec)
  # Note that parseOxcalOutput() defaults to only returning individual dates
  oxcal_data <- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)

@@

# The data is returned as a list of lists that needs to be unpacked
rcarb_data <- do.call(rbind, posteriors) 
  group_by(site) %>%
  mutate(meandate = mean(date_grid * -1)) %>% 
  ungroup() %>% 
  arrange(meandate)

h3 <- radiocarbon %>% 
  filter(site == "Vallermyrene 1" | site == "Stokke/Polland 5")

v1 <- model_rcarb(h3)

h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)

my_result_file <- executeOxcalScript(oxcal_code)
my_result_text <- readOxcalOutput(my_result_file)
my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = FALSE)
plot(my_result_data$`Vallermyrene 1`$posterior_probabilities$dates, my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities, type = "l")

posterior <- data.frame(probabilities = my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities,
                        date_grid = my_result_data$`Vallermyrene 1`$posterior_probabilities$dates,
                        site = "Vallermyrene 1",
                        excavation_project = "Vestfoldbaneprosjektet")

ggplot() +
  geom_line(aes(x = yrs, y = my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities))

yrs <- seq(11500, 5000, -500)
plt1 <- ggplot(sites, aes(x = yrs)) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
    scale_x_reverse()

ggplot(sites, aes(x = yrs)) +
      ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
                                                      y = fct_reorder(site, date_grid * -1, .fun = mean),
                                      height = probabilities * 100,
                                      fill = excavation_project)) +
    geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
    scale_x_reverse()

    ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
                                      y = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
                                      height = probabilities * 100,
                                      group = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
                                      fill = excavation_project)) +
  scale_fill_manual(labels = c("E18 Rugtvedt-Dørdal", 
                                "Vestfoldbaneprosjektet", "E18 Bommestad-Sky"),
                     values = c("springgreen4",
                                "gold1", "firebrick3"),
                    guide = FALSE) +

pst <- get_posterior_probabilities(my_result_data)
# h3_sum <- oxcal_Sum(h3dat)
# h3_run <- executeOxcalScript(h3_sum)
# h3_text <- readOxcalOutput(h3_run)
# h3_result <- parseFullOxcalOutput(h3_text)

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. The date ranges follow those given in the original excavation reports", fig.height = 10, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

yrs <- seq(11500, 5000, -500)
temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_x_reverse() +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))

@@

The date ranges provided for the sites are taken from the original excavation reports, and are the result of a consideration of typological indicators, dating with reference to shoreline displacement, and radiocarbon dates where these are available. Employing these date-ranges was deemed sufficient, as the main purpose is to get at general tendencies in the material, and not fine-grained chronological developments.

@@

While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context. This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is a higher number of cores and unretouched flakes and blades [@clark2017].

@@

corr_plot(lithicstr)
```

@@

pcalithics <- PCA(select(lithicspca, -mean_date),
                  scale.unit = TRUE, graph = FALSE)
fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithicspca$mean_date / 1000, 
          repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
         title = "") + 
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

>>>>

```{r libraries, eval = TRUE}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

@@

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.

However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

```{r c14-modelling, cache = TRUE}
# define function to model radiocarbon dates. Only assumption is that 
# dates from a single site belong to a single group
model_rcarb <- function(data){

@@

  # Run OxCal code - wrapped this in some functions to  not print to 
  # rmarkdown
  invisible(capture.output(oxcal_exec <- executeOxcalScript(oxcal_code)))
  oxcal_read <- readOxcalOutput(oxcal_exec)
  # Note that parseOxcalOutput() defaults to only returning individual dates,
  # specified with only.R_Date
  oxcal_data <- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)

@@

# The data is returned as a list of lists that needs to be unpacked
rcarb_data <- do.call(rbind, posteriors)

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon probability density functions are given as the posterior sum. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 11, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

# Combine radiocarbon data and reported (effectively shoreline/typology) dates,
# removing the reported date if there are relevant 14C-dates.
datedata <- full_join(sites, rcarb_data, by = "site") %>% 
  mutate(reported_start_bc = ifelse(dating_method != "c14", reported_start_bc, NA),
         reported_end_bc = ifelse(dating_method != "c14", reported_end_bc, NA))

project_colors <- c("bommestad_sky" = "springgreen4",
                  "brunlanes" = "royalblue2", 
                  "rugtvedt_dordal" = "gold1",
                  "vestfoldbaneprosjektet" = "firebrick3")

# Assemble plot of date ranges
temp_dist <- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = site, colour = excavation_project.x,
                   fill = excavation_project.x), size = 1.8) +
  geom_ridgeline(aes(x = date_grid * -1, height = probabilities * 150,
                     fill = excavation_project.y, 
                     colour = excavation_project.x)) +
  scale_x_reverse() +
  scale_fill_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  scale_colour_manual(values = project_colors) +
  labs(title = "B", fill = "Excavation project") +
  xlab("Cal. BC") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
         colour = FALSE)

@@

Where available radiocarbon dates associated with the sites have been subjected to bayesian modelling employing Oxcal through the oxcAAR package in R. The only constraint imposed in the model was that the dates from each site are assumed to belong to a group ---

title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"

author:

-   Isak Roalkvam:

email: isak.roalkvam\@iakh.uio.no

institute: [UiO]

correspondence: true

institute:

-   UiO: University of Oslo

date: "\`r format(Sys.time(), '%d %B, %Y')\`"

output:

bookdown::html_document2:

fig_caption: yes

reference_docx: "../templates/template.docx" \# Insert path for the DOCX file

pandoc_args:

-   --lua-filter=../templates/scholarly-metadata.lua

-   --lua-filter=../templates/author-info-blocks.lua

-   --lua-filter=../templates/pagebreak.lua

bibliography: references.bib

csl: "../templates/journal-of-archaeological-science.csl" \# Insert path for the bib-style

------------------------------------------------------------------------

\`\`\`{r setup, echo = FALSE}

knitr::opts_chunk\$set(

collapse = TRUE,

warning = FALSE,

message = FALSE,

echo = FALSE,

comment = "\#\>",

fig.width = 9,

fig.path = "../figures/",

dpi = 300

)

\`\`\`

\`\`\`{r libraries, eval = TRUE}

library(tidyverse)

library(GGally)

library(ggridges)

library(patchwork)

library(gt)

library(FactoMineR)

library(factoextra)

library(sf)

library(rnaturalearth)

library(oxcAAR)

quickSetupOxcal()

\`\`\`

\`\`\`{r get-functions, include = FALSE}

source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())

\#setOxcalExecutablePath("/tmp/RtmpFVOcBJ")

\`\`\`

\`\`\`{r get-data, eval = TRUE, results = 'hide'}

assemblages \<- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))

radiocarbon \<- read.csv(here::here('analysis/data/raw_data/radiocarbon_data.csv'))

site_pts \<- st_read(here::here('analysis/data/raw_data/sites.gpkg'))

\# ne_download() is currently failing

\#countries \<- ne_download(scale = 10, type = 'countries', category = 'physical',

\# returnclass = "sf")

countries \<- st_read(here::here('analysis/data/raw_data/naturalearth_countries.gpkg'))

\# gis \<- read.csv(here::here('../../site_data.csv'))

\# gis\$askeladden_id \<- as.integer(str_replace_all(gis\$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))

\# gisjoin \<- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))

\`\`\`

\# Introduction

The analyses conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. \@dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials and seasonality. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.

Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [\@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

\#\# Archaeological context and material

\`\`\`{r wrangling}

sites \<- assemblages %\>%

\# Find mean reported date for each site (to be used in the plots below)

rowwise() %\>%

mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %\>%

mutate(

\# Find total artefact count and flint count for each site

artefact_count = sum(c_across(core:flake_retouch_other)),

flint_count = sum(c_across(core:flake_retouch)),

nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %\>%

\# Exclude sites with less than 100 artefacts

filter(artefact_count \> 100)

\`\`\`

\`\`\`{r c14-modelling, cache = TRUE}

\# define function to model radiocarbon dates. Only assumption is that

\# dates from a single site belong to a single group

model_rcarb \<- function(data){

\# Format individual dates for R_Date in Oxcal

dates \<- R_Date(data[["lab_code"]], data[["c14_bp"]], data[["error"]])

\# Manually compile rest of OxCal code

oxcal_code \<- paste0('

Plot()

{

Sequence()

{

Boundary("Start");

Phase()

{

',

dates,

'

Sum("Sum");

};

Boundary("End");

};

};')

\# Run OxCal code - wrapped this in some functions to not print to

\# rmarkdown

invisible(capture.output(oxcal_exec \<- executeOxcalScript(oxcal_code)))

oxcal_read \<- readOxcalOutput(oxcal_exec)

\# Note that parseOxcalOutput() defaults to only returning individual dates,

\# specified with only.R_Date

oxcal_data \<- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)

\# Assemble data frame to be returned

posterior \<- data.frame(

probabilities = oxcal_data\$Sum\$posterior_probabilities\$probabilities,

date_grid = oxcal_data\$Sum\$posterior_probabilities\$dates,

site = data[["site"]][1],

excavation_project = data[["excavation_project"]][1])

\# Return data

return(posterior)

}

\# Apply modelling function to the radiocarbon dates, grouped by site name

posteriors \<- radiocarbon %\>%

group_by(site) %\>%

group_map(\~ model_rcarb(.), .keep = TRUE)

\# The data is returned as a list of lists that needs to be unpacked

rcarb_data \<- do.call(rbind, posteriors)

\`\`\`

The \`r nrow(sites)\` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \\ref\\\@(fig:dist)A]. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [\@solheim2013b; \@jaksland2014a; \@melvold2014b; \@reitan2014l; \@solheim2017b]. The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. \@clark2006; \@dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [\@solheim2013b; \@jaksland2014, see also \\ref\\\@(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

\`\`\`{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon probability density functions are given as the posterior sum. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 12, fig.width = 12}

\# Create a bounding box around site points to limit the overview map

sitbbox \<- st_bbox(site_pts)

sitbbox[1:2] \<- sitbbox[1:2] - 600000

sitbbox[3:4] \<- sitbbox[3:4] + 600000

boundingpoly \<- st_as_sf(st_as_sfc(sitbbox))

\# Reproject the bounding box to match world map, and crop the world map

\# with the bounding box polygon

bound_reproj \<- st_transform(boundingpoly, st_crs(countries))

mapcountries \<- countries %\>%

filter(st_intersects(., bound_reproj, sparse = FALSE))

count_reproj \<- st_transform(mapcountries, st_crs(site_pts))

\# Create bounding box around the sites, both for extent indicator on the

\# overview map and to create basemap for the site overview.

sitbbox2 \<- st_bbox(site_pts)

sitbbox2[1:2] \<- sitbbox2[1:2] - 15000

sitbbox2[3:4] \<- sitbbox2[3:4] + 15000

boundingpoly2 \<- st_as_sf(st_as_sfc(sitbbox2))

\# same as above, using the smaller bounding box

bound_reproj2 \<- st_transform(boundingpoly2, st_crs(countries))

mapcountries2 \<- countries %\>%

filter(st_intersects(., bound_reproj2, sparse = FALSE))

count_reproj2 \<- st_transform(mapcountries2, st_crs(site_pts))

\# Assemble overview map

overview \<-

ggplot() +

geom_sf(data = count_reproj, fill = "grey", colour = NA) +

geom_sf(data = boundingpoly2,

fill = NA, colour = "black", size = 1) +

coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),

expand = FALSE) +

labs(title = "A") +

theme_bw() + theme(axis.title=element_blank(),

axis.text.y = element_blank(),

axis.text.x = element_blank(),

rect = element_rect(),

axis.ticks = element_blank(),

panel.grid.major = element_blank())

\# Combine spatial point data with assemblage data (holds excavation project

\# for colouring points in the maps)

sitejoin \<- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))

\# Assemble site map. anc defined first is used for placing the scalebar,

\# which required some tweaking to get right.

anc \<- as.numeric(c(sitbbox2\$ymin, sitbbox2\$xmax))

sitemap \<-

ggplot() +

geom_sf(data = count_reproj2, fill = "grey", colour = NA) +

geom_sf(data = sitejoin, aes(fill = excavation_project),

size = 2.25, shape=21,

colour = "black") +

labs(fill = "Excavation project") +

scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",

"E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),

values = c("springgreen4", "royalblue2",

"gold1", "firebrick3")) +

ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km",

transform = FALSE, st.size = 4, height = 0.05,

border.size = 0.05, st.dist = 0.07,

anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +

coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]),

ylim = c(sitbbox2[2], sitbbox2[4]),

expand = FALSE) +

theme_bw() + theme(axis.title=element_blank(),

axis.text.y = element_blank(),

axis.text.x = element_blank(),

rect = element_rect(),

axis.ticks = element_blank(),

panel.grid.major = element_blank(),

legend.position = "none") +

guides(fill = guide_legend(ncol = 2,

title.position = "top", title.hjust = 0.5))

\# Combine radiocarbon data and reported (effectively shoreline/typology) dates,

\# removing the reported date if there are relevant 14C-dates.

datedata \<- full_join(sites, rcarb_data, by = "site") %\>%

mutate(reported_start_bc = ifelse(dating_method != "c14", reported_start_bc, NA),

reported_end_bc = ifelse(dating_method != "c14", reported_end_bc, NA))

project_colors \<- c("bommestad_sky" = "springgreen4",

"brunlanes" = "royalblue2",

"rugtvedt_dordal" = "gold1",

"vestfoldbaneprosjektet" = "firebrick3")

\# Assemble plot of date ranges

temp_dist \<- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +

geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,

yend = site, colour = excavation_project.x,

fill = excavation_project.x), size = 1.8) +

geom_ridgeline(aes(x = date_grid \* -1, height = probabilities \* 150,

fill = excavation_project.y,

colour = excavation_project.x)) +

scale_x\_reverse() +

scale_fill_manual(values = project_colors,

labels = c("E18 Bommestad-Sky","E18 Brunlanes",

"E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +

scale_colour_manual(values = project_colors) +

labs(title = "B", fill = "Excavation project") +

xlab("Cal. BC") +

theme_classic() +

theme(axis.title.y=element_blank(),

axis.text.y=element_blank(),

axis.ticks.y=element_blank(),

legend.position = "bottom") +

guides(fill = guide_legend(title.position = "top", title.hjust = 0.5),

colour = FALSE)

(overview + sitemap) /

(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))

\`\`\`

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [\@glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [\@breivik2018; \@solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. \@sørensen1979; \@romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. \@perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of \@bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [\@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [\@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

Where available, radiocarbon dates associated with the sites have been subjected to bayesian modelling employing OxCal through the oxcAAR package in R. The only constraint imposed in the models was that the dates from each site are assumed to belong to the same group employing the Phase function from OxCal. The posterior distributions were then summed for each site. Radiocarbon dates considered irrelevant or erroneous in the original reports were disregarded. Where relevant radiocarbon dates are not available, the date ranges provided for the sites are taken from the original excavation reports. These are the result of a consideration of typological indicators as well as shoreline displacement and local topography.

The assemblage data analysed here is based on the classification of the \`r nrow(sites)\` site assemblages done for the original excavation reports, and consists of \`r ncol(select(sites, core:flake_retouch_other))\` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width \$\\ge\$ 12 mm) and micro-blades (width \$\\le\$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. \@berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [\@eigeland2015, 46]. An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites. However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [\@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. \@frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [\@eigeland2015, 370].

@@

While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context. This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is as a consequence also a higher number of cores and unretouched flakes and blades [@clark2017].

@@

corr_plot(select(lithicstr, -reported_start_bc, -reported_end_bc, -dating_method))
```

@@

pcalithics <- PCA(select(lithicspca, -reported_start_bc, -dating_method,
                         -reported_end_bc, -mean_date),
                  scale.unit = TRUE, graph = FALSE)
fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithicspca$mean_date / 1000, 
          repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
         title = "") + 
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")
